---
title: "pt2_station_avg_regressions_CHLa_TN_TP"
author: "Andrew Cameron"
date: "2024-11-24"
output: html_document
---

Derive  average values (May-Sep) for each year and array results by column.										
Note that most stations were sampled in only 1 of 2 years.										
Note that CHLa, TN and TP were not all measured at every station.			

Fields:
StationID	StationName	AdminRegion	Depth	CHLa23	TN23	TP23	CHLa24	TN24	TP24	

```{r}
library(tidyverse)
data <- read_csv("data/step1_data_MaytoSep_wide.csv")

other_columns <- data %>% 
  select(StationId, StationName, AdminRegion)   # to be joined back later. Can't be retained in grouping/summarizing because lots of NA for StationName and AdminRegion.

#Derive average values for each year
data_avg <- data %>%
  mutate(year = year(DateTime)) %>%
  group_by(StationId, Depth, year) %>%
  summarise(CHLa = mean(CHLa, na.rm = TRUE),
            TN = mean(TN, na.rm = TRUE),
            TP = mean(TP, na.rm = TRUE),
            .groups = "drop")

#Array results by column
data_avg_wide <- data_avg %>%
  pivot_wider(names_from = year, values_from = c(CHLa, TN, TP))

desired_names <- c("CHLa23", "CHLa24", "TN23", "TN24", "TP23", "TP24")
names(data_avg_wide)[3:8] <- desired_names

#Bring back StationName and AdminRegion
stationName <- other_columns %>%
  group_by(StationId) %>%
  summarise(
    StationName = coalesce(first(StationName[!is.na(StationName)]), first(StationName))
  ) %>%
  ungroup()

adminRegion <- other_columns %>%
  group_by(StationId) %>%
  summarise(
    AdminRegion = coalesce(first(AdminRegion[!is.na(AdminRegion)]), first(AdminRegion))
  ) %>%
  ungroup()

data_avg_wide <- data_avg_wide %>%
  left_join(stationName, by = "StationId") %>%
  left_join(adminRegion, by = "StationId") %>%
  relocate(StationName, AdminRegion, .before = Depth)

write_csv(data_avg_wide, "data/step2_station_avg_by_year.csv")

```

## Output 1: Regressions of station average values for 2023 vs. 2024

How many stations are actually sampled in both years?

```{r}
both_chla <- data_avg_wide %>%
  filter(!is.na(CHLa23) & !is.na(CHLa24))

both_tn <- data_avg_wide %>%
  filter(!is.na(TN23) & !is.na(TN24))

both_tp <- data_avg_wide %>%
  filter(!is.na(TP23) & !is.na(TP24))

print(paste("CHLa: ", nrow(both_chla)))
print(paste("TN: ", nrow(both_tn)))
print(paste("TP: ", nrow(both_tp)))

```

```{r, warning = FALSE}

modelSummaries <- list()  
  
  for (var in c("CHLa", "TN", "TP")) {
  
    model_data <- data_avg_wide %>%
      filter(!is.na(!!sym(paste0(var, "23"))) & !is.na(!!sym(paste0(var, "24"))))
    
    modeling_cols <- c(paste0(var, "23"), paste0(var, "24"))
    formula <- as.formula(paste0(modeling_cols[2], " ~ ", modeling_cols[1]))
    
    # Run mblm model
    model <- mblm::mblm(formula, data = model_data)
    mod.sum <- mblm::summary.mblm(model)
    
    # Store results (directly, without creating the combined station-variable name)
    modelSummary <- list(
      variable = var,
      slope = mod.sum$coefficients[2, 1],
      MAD = mod.sum$coefficients[modeling_cols[1], "MAD"],
      pvalue = mod.sum$coefficients[modeling_cols[1], 4],
      intercept = mod.sum$coefficients[1, 1]
    )
    
    modelSummaries[[var]] <- modelSummary
  }
  
```

Extract Regression Statistics into single table

```{r `regression stats`}
 reg.df <- data.frame(variable = c("CHLa", "TN", "TP"),
               model_slope = rep(NA, 3),
               model_MAD = rep(NA, 3),
               model_pval = rep(NA, 3),
               model_intercept = rep(NA, 3))

for (i in seq_along(1:3)) {
  var <- c("CHLa", "TN", "TP")[i]
  modelSum <- modelSummaries[[var]]
      
        reg.df$model_slope[i] <- modelSum$slope
        reg.df$model_MAD[i] <- modelSum$MAD
        reg.df$model_pval[i] <-  modelSum$pvalue
        reg.df$model_intercept[i] <-  modelSum$intercept
      }


write_csv(reg.df, "data/step2_2023vs2024.csv")

#Prepare regression df for plotting
reg_plot <- reg.df %>%
  mutate(pval_render = case_when(
    model_pval > 0.05 ~ paste0("p = ", formatC(model_pval, digits = 2, format = "f")),
    model_pval < 0.05 & model_pval >= 0.001 ~ paste0("p = ", formatC(model_pval, digits = 3, format = "f")),
    model_pval < 0.001 ~ "p < 0.001"
  ),
  y_label = case_when(
    variable == "CHLa" ~ "CHLa (\u03BCg L\u207B\u00B9)",
    variable == "TP" ~ "TP (mg L\u207B\u00B9)",
    variable == "TN" ~ "TN (mg L\u207B\u00B9)"
  ))

```

Data viz

```{r}
# !------------------- CHLa -------------------------! 
data_chla <- data_avg_wide %>%
      filter(!is.na(CHLa23) & !is.na(CHLa24))
reg_chla <- reg_plot[reg_plot$variable == "CHLa",]
max_y <- max(data_chla$CHLa23, na.rm = TRUE) * 1.2
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(data_chla$CHLa23) - 12  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.95           # Place it slightly below the upper limit of the y-axis

ggplot(data_chla, aes(x = CHLa23, y = CHLa24)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = reg_chla$pval_render, size = 4.2, hjust = 1) +
      # Regression line
      annotate("segment", x = min(data_chla$CHLa23), 
               xend = max(data_chla$CHLa23), 
               y = reg_chla$model_intercept , 
               yend = reg_chla$model_intercept  + reg_chla$model_slope * (max(data_chla$CHLa23)),
               color = "red", linewidth = .42, alpha = .6) +
      labs(title = NULL,
           x ="CHLa (\u03BCg L\u207B\u00B9) 2023", 
           y = "CHLa (\u03BCg L\u207B\u00B9) 2024") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )
ggsave("figures/CHLa_23vs24.jpg", width = 8, height = 6, dpi = 500)

# !------------------- TN -------------------------! 
data_TN <- data_avg_wide %>%
      filter(!is.na(TN23) & !is.na(TN24))
reg_TN <- reg_plot[reg_plot$variable == "TN",]
max_y <- max(data_TN$TN23, na.rm = TRUE) * 1.2
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(data_TN$TN23) -.05  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.5           # Place it slightly below the upper limit of the y-axis

ggplot(data_TN, aes(x = TN23, y = TN24)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = reg_TN$pval_render, size = 4.2, hjust = 1) +
      # Regression line
      annotate("segment", x = min(data_TN$TN23), 
               xend = max(data_TN$TN23), 
               y = reg_TN$model_intercept , 
               yend = reg_TN$model_intercept  + reg_TN$model_slope * (max(data_TN$TN23)),
               color = "red", linewidth = .42, alpha = .6) +
      labs(title = NULL,
           x ="TN (\u03BCg L\u207B\u00B9) 2023", 
           y = "TN (\u03BCg L\u207B\u00B9) 2024") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )
ggsave("figures/TN_23vs24.jpg", width = 8, height = 6, dpi = 500)

# !------------------- TP -------------------------! THIS ONE IS FUCKED UP ???
data_TP <- data_avg_wide %>%
      filter(!is.na(TP23) & !is.na(TP24))
reg_TP <- reg_plot[reg_plot$variable == "TP",]
max_y <- max(data_TP$TP23, na.rm = TRUE) * 1.2
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(data_TP$TP23) -.05  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.5           # Place it slightly below the upper limit of the y-axis

ggplot(data_TP, aes(x = TP23, y = TP24)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = reg_TP$pval_render, size = 3.2, hjust = 1) +
      # Regression line
      annotate("segment", x = min(data_TP$TP23), 
               xend = max(data_TP$TP23), 
               y = reg_TP$model_intercept , 
               yend = reg_TP$model_intercept  + reg_TP$model_slope * (max(data_TP$TP23)),
               color = "red", linewidth = .42, alpha = .6) +
      labs(title = NULL,
           x ="TP (\u03BCg L\u207B\u00B9) 2023", 
           y = "TP (\u03BCg L\u207B\u00B9) 2024") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 9.5),        # Adjust axis title size if needed
        axis.text = element_text(size = 7.5, color = "black"),         # Adjust axis text size
        plot.subtitle = element_text(size = 9),     # Adjust subtitle size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )

```

Output 2: regressions of station average values for CHLa vs. TP, CHLa vs. TN and TP vs. TN (stations samples in both years appear as 2 data points).	

```{r}

```