---
title: 'Part 3: Annual Means vs. Long-Term Means by Variable'
output: html_document
---

*Merge Annual average values for 2023 and 2024 with previously derived long-term average values (LTMs).	*	

Note that this survey only includes a small subset of all the stations for which we have derived long-term average values.				
Note that this survey also includes new stations for which there are no long-term average values.		

Output: regressions of annual means vs. long-term means (e.g., CHLa23 and CHLa24 vs. CHLaLTM).	

Fields:
StationID	StationName	AdminRegion	Depth	CHLa23	TN23	TP23	CHLa24	TN24	TP24	CHLaLTM	TNLTM	TPLTM

# ---------------------

Our previous LTMs are based on yearly averages for May-Oct, whereas Paul has specified that the months of interest here are May-Sep. 
Will need to derive new yearly and LTMs for previous dataset before incorporating the new survey data.
?? Ensure same methodology for yearly averages -- 'first' or 'mean' when multiple measurements on same date?

Steps
bring in previous water quality data 
filter to only include stations of interest (i.e., those that are both in this survey and have LTM from previous analysis)
merge new data (?) -- but isn't the "new" data already in the old data, at least the 2023, if from the same station?
derive LTM from combined data
generate output csv
regressions
plots

## Load and Prepare Data
```{r}
library(tidyverse)
# Data from previous water clarity analysis
prev_data <- read_csv("~/Desktop/Bukaveckas_WaterLab/water-clarity-analysis/data/processed_dataset.csv") %>%
  select(STATION_ID, DATE, CHLa, NITROGEN_TOTAL, PHOSPHORUS_TOTAL) %>%
  rename(TN = NITROGEN_TOTAL, TP = PHOSPHORUS_TOTAL)

prev_LTMs <- read_csv("~/Desktop/Bukaveckas_WaterLab/water-clarity-analysis/data/station_level_data/stationAverages_longTerm.csv")
cur_data <- read_csv("data/StationObs_MaytoSep_wide.csv")

# Stations of focus for this regression
index <- which(unique(cur_data$StationId) %in% prev_LTMs$STATION_ID)
stations <- unique(cur_data$StationId)[index]   # only 4 stations in common


# Subset data from each data set to only include stations of interest
subset_current <- cur_data %>%
  filter(StationId %in% stations) %>%
  mutate(dataset = "CurrentAnalysis") %>%
  select(-c(AdminRegion, StationName, Depth)) %>%
  rename(DATE = DateTime,
         STATION_ID = StationId)  # to match column names

subset_prev <- prev_data %>%
  filter(STATION_ID %in% stations)  %>%
  filter(month(DATE) != 10) %>%
  mutate(dataset = "PreviousAnalysis") %>%
  rename(DATE = DATE)  # to match column names

# Merge data sets
combined_data <- rbind(subset_current, subset_prev) %>%
  arrange(STATION_ID, DATE)

# Take mean of multiple observations on same date
combined_data <- combined_data %>%
  group_by(STATION_ID, DATE) %>%
  summarize(CHLa = mean(CHLa, na.rm = TRUE),
            TN = mean(TN, na.rm = TRUE),
            TP = mean(TP, na.rm = TRUE),
            .groups = "drop")

```

## Annual Averages and LTMs

Derive annual averages by variable.
Then use those to calculate a long-term mean for each parameter.

```{r}
LTMs_list <- list()

for (var in c("CHLa", "TN", "TP")) {
  annual_var <- combined_data %>%
    group_by(STATION_ID, year = year(DATE)) %>%
    summarize(mean = mean(!!sym(var), na.rm = TRUE),
              .groups = "drop")
  
  LTM_var <- annual_var %>%
    group_by(STATION_ID) %>%
    summarize(LTM = mean(mean, na.rm = TRUE))
  
  LTMs_list[[paste0(var, "LTM")]] <- LTM_var

}

LTMs <- bind_rows(LTMs_list, .id = "param") %>%
  pivot_wider(names_from = param, values_from = LTM)
## 2-APP020.23 did NOT meet 10y threshold for inclusion in LTMs and trend statistics during water clarity analysis for either CHLa or TN. Only for TP


# Join averages and LTMs
stationAvgs <- read_csv("data/station_avg_by_year.csv")
ann_plus_LTM <- stationAvgs %>%
  left_join(LTMs, by = c("StationId" = "STATION_ID"))


```

## Regression Analyis: Annual Means vs. Long-Term Means

```{r}
# Prepare regression data
## CHLa
chla_data <- ann_plus_LTM %>%
  filter(StationId %in% stations) %>%
  select(StationId, CHLa23, CHLa24) %>%
  gather(key = "variable", value = "value", -StationId) %>%
  mutate(variable = factor(variable, levels = c("CHLa23", "CHLa24"))) %>%
  left_join(LTMs, by = c("StationId" = "STATION_ID"))

## TN 
tn_data <- ann_plus_LTM %>%
  filter(StationId %in% stations) %>%
  select(StationId, TN23, TN24) %>%
  gather(key = "variable", value = "value", -StationId) %>%
  mutate(variable = factor(variable, levels = c("TN23", "TN24"))) %>%
  left_join(LTMs, by = c("StationId" = "STATION_ID"))

## TP
tp_data <- ann_plus_LTM %>%
  filter(StationId %in% stations) %>%
  select(StationId, TP23, TP24) %>%
  gather(key = "variable", value = "value", -StationId) %>%
  mutate(variable = factor(variable, levels = c("TP23", "TP24"))) %>%
  left_join(LTMs, by = c("StationId" = "STATION_ID"))

reg_data <- rbind(chla_data, tn_data, tp_data)

```

```{r}
# Run regressions
modelSummaries <- list()  
  
  for (var in c("CHLa", "TN", "TP")) {
    
    model_data <- reg_data %>%
      filter(grepl(var, variable)) %>%
      filter(!is.na(value))
    
    x_col <- paste0(var, "LTM")
    formula <- as.formula(paste("value ~", x_col))
    
    # Run mblm model
    model <- mblm::mblm(formula, data = model_data)
    mod.sum <- mblm::summary.mblm(model)
    
    # Store results (directly, without creating the combined station-variable name)
    modelSummary <- list(
      variable = var,
      slope = mod.sum$coefficients[2, 1],
      MAD = mod.sum$coefficients[x_col[1], "MAD"],
      pvalue = mod.sum$coefficients[x_col[1], 4],
      intercept = mod.sum$coefficients[1, 1]
    )
    
    modelSummaries[[var]] <- modelSummary
  }

```

#### Extract Regression Statistics into single table

```{r `regression stats`}
 reg.df <- data.frame(variable = c("CHLa", "TN", "TP"),
               model_slope = rep(NA, 3),
               model_MAD = rep(NA, 3),
               model_pval = rep(NA, 3),
               model_intercept = rep(NA, 3))

for (i in seq_along(1:3)) {
  var <- c("CHLa", "TN", "TP")[i]
  modelSum <- modelSummaries[[var]]
      
        reg.df$model_slope[i] <- modelSum$slope
        reg.df$model_MAD[i] <- modelSum$MAD
        reg.df$model_pval[i] <-  modelSum$pvalue
        reg.df$model_intercept[i] <-  modelSum$intercept
      }

#Prepare regression df for plotting
reg_plot <- reg.df %>%
  mutate(pval_render = case_when(
    model_pval > 0.05 ~ paste0("p = ", formatC(model_pval, digits = 2, format = "f")),
    model_pval < 0.05 & model_pval >= 0.001 ~ paste0("p = ", formatC(model_pval, digits = 3, format = "f")),
    model_pval < 0.001 ~ "p < 0.001"
  ),
  y_label = case_when(
    variable == "CHLa" ~ "CHLa (\u03BCg L\u207B\u00B9)",
    variable == "TP" ~ "TP (mg L\u207B\u00B9)",
    variable == "TN" ~ "TN (mg L\u207B\u00B9)"
  ))

```

#### Merge Annual average (2023 and 2024) and long-term average values with staiton-specific trend statistics.

The newly derived annual averages and long-term averages should be merged with the station-specific trend statistics from previous water clarity analysis.
"Note that only a small proportion (possibly none) of the microcystin survey sites will have staiton-specific trend estimates for CHLa, TN or TP."

In the output dataframe, one station - 2-APP020.23 - has no station-specific trend statistics for CHLa or TN, despite having a long-term mean. Checking the metadata reveals that for CHLa, the station had *9* years of data, where our threshold was *10* for inclusion in the trend analysis. 

```{r}
station_specific_trends <- read_csv("~/Desktop/Bukaveckas_WaterLab/water-clarity-analysis/data/station_level_data/longTerm_Annual_Regression_Statistics_ALLSTATIONS.csv") %>%
    select(-model_intercept) %>%
    mutate(variable = case_when(
      grepl("NITROGEN", variable) ~ "TN",
      grepl("PHOSPHORUS", variable) ~ "TP",
      TRUE ~ variable
    )) %>%
    filter(variable %in% c("CHLa", "TN", "TP"))


x <- station_specific_trends %>%
  pivot_wider(names_from = variable, values_from = c(model_slope, model_MAD, model_pval)) %>%
  select(
    STATION_ID,
    contains("CHLa"),
    contains("TN"),
    contains("TP"),
    everything()
  )

colnames(x) <- gsub("model_", "", colnames(x))


# Merge with annual averages and LTMs
output <- ann_plus_LTM %>%
  left_join(x, by = c("StationId" = "STATION_ID"))

write_csv(output, "data/final_processed_dataset.csv")


```


#### Data viz

```{r}
# !------------------- CHLa -------------------------! 
data_chla <- reg_data %>%
      filter(grepl("CHLa", variable)) %>%
      filter(!is.na(value))
    
reg_chla <- reg_plot[reg_plot$variable == "CHLa",]
max_y <- max(data_chla$value, na.rm = TRUE) * 1.2
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(data_chla$CHLaLTM) - .5  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.8           # Place it slightly below the upper limit of the y-axis

ggplot(data_chla, aes(x = CHLaLTM, y = value)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = reg_chla$pval_render, size = 4.2, hjust = 1) +
      # Regression line
      annotate("segment", x = min(data_chla$CHLaLTM), 
               xend = max(data_chla$CHLaLTM), 
               y = reg_chla$model_intercept + reg_chla$model_slope * min(data_chla$CHLaLTM), 
               yend = reg_chla$model_intercept  + reg_chla$model_slope * (max(data_chla$CHLaLTM)),
               color = "red", linewidth = .42, alpha = .6) +
      labs(title = NULL,
           x ="CHLa (\u03BCg L\u207B\u00B9) LTM", 
           y = "CHLa (\u03BCg L\u207B\u00B9) May-Sep Annual Mean") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )
ggsave("figures/CHLa_annualVsLTM.jpg", width = 8, height = 6, dpi = 500)

# !------------------- TN -------------------------! 
data_TN <- reg_data %>%
      filter(grepl("TN", variable)) %>%
      filter(!is.na(value))
    
reg_TN <- reg_plot[reg_plot$variable == "TN",]
max_y <- max(data_TN$value, na.rm = TRUE) * 1.2
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(data_TN$TNLTM)  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.8           # Place it slightly below the upper limit of the y-axis

ggplot(data_TN, aes(x = TNLTM, y = value)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = reg_TN$pval_render, size = 4.2, hjust = 1) +
      # Regression line
      annotate("segment", x = min(data_TN$TNLTM), 
               xend = max(data_TN$TNLTM), 
               y = reg_TN$model_intercept + reg_TN$model_slope * min(data_TN$TNLTM), 
               yend = reg_TN$model_intercept  + reg_TN$model_slope * (max(data_TN$TNLTM)),
               color = "red", linewidth = .42, alpha = .6) +
      labs(title = NULL,
           x ="TN (\u03BCg L\u207B\u00B9) LTM", 
           y = "TN (\u03BCg L\u207B\u00B9) May-Sep Annual Mean") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )
ggsave("figures/TN_annualVsLTM.jpg", width = 8, height = 6, dpi = 500)


# !------------------- TP -------------------------!
data_TP <- reg_data %>%
      filter(grepl("TP", variable)) %>%
      filter(!is.na(value))
    
reg_TP <- reg_plot[reg_plot$variable == "TP",]
max_y <- max(data_TP$value, na.rm = TRUE) * 1.2
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(data_TP$TPLTM)   # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.8           # Place it slightly below the upper limit of the y-axis

ggplot(data_TP, aes(x = TPLTM, y = value)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = reg_TP$pval_render, size = 4.2, hjust = 1) +
      # Regression line
      annotate("segment", x = min(data_TP$TPLTM), 
               xend = max(data_TP$TPLTM), 
               y = reg_TP$model_intercept + reg_TP$model_slope * min(data_TP$TPLTM), 
               yend = reg_TP$model_intercept  + reg_TP$model_slope * (max(data_TP$TPLTM)),
               color = "red", linewidth = .42, alpha = .6) +
      labs(title = NULL,
           x ="TP (\u03BCg L\u207B\u00B9) LTM", 
           y = "TP (\u03BCg L\u207B\u00B9) May-Sep Annual Mean") +
  scale_y_continuous(breaks = seq(0.022, .033, 0.002)) +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )
ggsave("figures/TP_annualVsLTM.jpg", width = 8, height = 6, dpi = 500)

```
