---
title: 'Part 3: Annual Means vs. Long-Term Means by Variable'
output: html_document
---

*Merge Annual average values for 2023 and 2024 with previously derived long-term average values (LTMs).	*	

Note that this survey only includes a small subset of all the stations for which we have derived long-term average values.				
Note that this survey also includes new stations for which there are no long-term average values.		

Output: regressions of annual means vs. long-term means (e.g., CHLa23 and CHLa24 vs. CHLaLTM).	

Fields:
StationID	StationName	AdminRegion	Depth	CHLa23	TN23	TP23	CHLa24	TN24	TP24	CHLaLTM	TNLTM	TPLTM

The cleanest result would be regressions of 2023 means vs. LTMs based on all preceding years (e.g., up to 2022) and regressions of 2024 means vs. LTMs of all preceding years (up to 2023).  
That way we are directly testing how good of a predictor is the long-term mean of a future annual mean.  

# ---------------------

Our previous LTMs are based on yearly averages for May-Oct, whereas Paul has specified that the months of interest here are May-Sep. 
Will need to derive new yearly and LTMs for previous dataset before incorporating the new survey data.
?? Ensure same methodology for yearly averages -- 'first' or 'mean' when multiple measurements on same date?

Steps
bring in previous water quality data 
filter to only include stations of interest (i.e., those that are both in this survey and have LTM from previous analysis)
merge new data (?) -- but isn't the "new" data already in the old data, at least the 2023, if from the same station?
derive LTM from combined data
generate output csv
regressions
plots

## Load and Prepare Data
```{r}
library(tidyverse)
# Data from previous water clarity analysis
prev_data <- read_csv("~/Desktop/Bukaveckas_WaterLab/water-clarity-analysis/data/processed_dataset.csv") %>%
  select(STATION_ID, DATE, CHLa, NITROGEN_TOTAL, PHOSPHORUS_TOTAL) %>%
  rename(TN = NITROGEN_TOTAL, TP = PHOSPHORUS_TOTAL)

prev_LTMs <- read_csv("~/Desktop/Bukaveckas_WaterLab/water-clarity-analysis/data/station_level_data/5y_summaryStats.csv")
cur_data <- read_csv("data/StationObs_MaytoSep_wide.csv")

# Stations of focus for this regression
index <- which(unique(cur_data$StationId) %in% prev_LTMs$STATION_ID)
stations <- unique(cur_data$StationId)[index]  

# Subset data from each data set to only include stations of interest
subset_current <- cur_data %>%
  filter(StationId %in% stations) %>%
  mutate(dataset = "CurrentAnalysis") %>%
  select(-c(AdminRegion, StationName, Depth, year_group)) %>%
  rename(DATE = DateTime,
         STATION_ID = StationId)  # to match column names

subset_prev <- prev_data %>%
  filter(STATION_ID %in% stations)  %>%
  filter(month(DATE) != 10) %>%
  mutate(dataset = "PreviousAnalysis") %>%
  rename(DATE = DATE)  # to match column names

# Merge data sets
combined_data <- rbind(subset_current, subset_prev) %>%
  arrange(STATION_ID, DATE)

# Take mean of multiple observations on same date
combined_data <- combined_data %>%
  group_by(STATION_ID, DATE) %>%
  summarize(CHLa = mean(CHLa, na.rm = TRUE),
            TN = mean(TN, na.rm = TRUE),
            TP = mean(TP, na.rm = TRUE),
            .groups = "drop")

```


## Annual Averages and LTMs

Derive annual averages by variable.
Then use those to calculate a long-term mean for each parameter.

Two LTMs are needed, one with data through 2022 (to compare with 2023) and one with data through 2023 (to compare with 2024).

```{r}
# LTMs with data through end of 2022 (i.e., "as of" 2023)
LTMs_list_2023 <- list()

for (var in c("CHLa", "TN", "TP")) {
  annual_var <- combined_data %>%
    filter(DATE < "2023-01-01") %>%
    group_by(STATION_ID, year = year(DATE)) %>%
    summarize(mean = mean(!!sym(var), na.rm = TRUE),
              .groups = "drop")
  
  LTM_var <- annual_var %>%
    group_by(STATION_ID) %>%
    summarize(LTM = mean(mean, na.rm = TRUE))
  
  LTMs_list_2023[[paste0(var, "LTM_2023")]] <- LTM_var

}

LTMs_2023 <- bind_rows(LTMs_list_2023, .id = "param") %>%
  pivot_wider(names_from = param, values_from = LTM)
## 2-APP020.23 did NOT meet 10y threshold for inclusion in LTMs and trend statistics during water clarity analysis for either CHLa or TN. Only for TP

# LTMs with data through end of 2023
LTMs_list_2024 <- list()

for (var in c("CHLa", "TN", "TP")) {
  annual_var <- combined_data %>%
    filter(DATE < "2024-01-01") %>%
    group_by(STATION_ID, year = year(DATE)) %>%
    summarize(mean = mean(!!sym(var), na.rm = TRUE),
              .groups = "drop")
  
  LTM_var <- annual_var %>%
    group_by(STATION_ID) %>%
    summarize(LTM = mean(mean, na.rm = TRUE))
  
  LTMs_list_2024[[paste0(var, "LTM_2024")]] <- LTM_var

}

LTMs_2024 <- bind_rows(LTMs_list_2024, .id = "param") %>%
  pivot_wider(names_from = param, values_from = LTM)

# Join LTMs and Station Avgs together
stationAvgs <- read_csv("data/station_avg_by_year.csv")

main_df <- stationAvgs %>%
  left_join(LTMs_2023, by = c("StationId" = "STATION_ID")) %>%
  left_join(LTMs_2024, by = c("StationId" = "STATION_ID"))

```

## Regression Analyis: Annual Means vs. Long-Term Means

```{r}
# Run regressions
# 2023 ~ LTM_thru_2022
modelSummaries_2023 <- list()  
  
  for (var in c("CHLa", "TN", "TP")) {
    predictor <- paste0(var,"LTM_2023")
    outcome <- paste0(var, "23")
    
    model_data <- main_df %>%
      filter(!is.na(!!sym(predictor)) & !is.na(!!sym(outcome))) %>%
      select(StationId, StationName, !!sym(predictor), !!sym(outcome))
    
    formula <- as.formula(paste(outcome, "~", predictor))
    
    # Run mblm model
    model <- lm(formula, data = model_data)
    mod.sum <- summary(model)
    
    # Store results (directly, without creating the combined station-variable name)
      modelSummary <- list(
        variable = var,
        slope = mod.sum$coefficients[2, 1],
        SE = mod.sum$coefficients[2, 2],
        intercept = mod.sum$coefficients[1, 1],
        R2 = mod.sum$r.squared,
        residualSE = mod.sum$sigma,
        pvalue = mod.sum$coefficients[2, 4]
      )
    
    modelSummaries_2023[[paste0(outcome, "~", predictor)]] <- modelSummary
  }

# -------------------------------------------------------

# 2024 ~ LTM_thru_2023
modelSummaries_2024 <- list()  
  
  for (var in c("CHLa", "TN", "TP")) {
    predictor <- paste0(var,"LTM_2024")
    outcome <- paste0(var, "24")
    
    model_data <- main_df %>%
      filter(!is.na(!!sym(predictor)) & !is.na(!!sym(outcome))) %>%
      select(StationId, StationName, !!sym(predictor), !!sym(outcome))
    
    formula <- as.formula(paste(outcome, "~", predictor))
    
    # Run mblm model
    model <- lm(formula, data = model_data)
    mod.sum <- summary(model)
    
    # Store results (directly, without creating the combined station-variable name)
      modelSummary <- list(
        variable = var,
        slope = mod.sum$coefficients[2, 1],
        SE = mod.sum$coefficients[2, 2],
        intercept = mod.sum$coefficients[1, 1],
        R2 = mod.sum$r.squared,
        residualSE = mod.sum$sigma,
        pvalue = mod.sum$coefficients[2, 4]
      )
    
    modelSummaries_2024[[paste0(outcome, "~", predictor)]] <- modelSummary
  }

```

### Extract Regression Statistics into single table

```{r `regression stats`}
# Extract Regression Statistics into Single Table
# 2023
reg.df_2023 <- data.frame(
  variable = c("CHLa", "TN", "TP"),
  model_slope = rep(NA, 3),
  model_SE = rep(NA, 3),
  model_pval = rep(NA, 3),
  model_intercept = rep(NA, 3),
  model_R2 = rep(NA, 3),
  residual_SE = rep(NA, 3)
)

for (i in seq_along(1:3)) {
  var <- c("CHLa", "TN", "TP")[i]
  predictor <- paste0(var,"LTM_2023")
  outcome <- paste0(var, "23")
  modelSum <- modelSummaries_2023[[paste0(outcome, "~", predictor)]]
      
  reg.df_2023$model_slope[i] <- modelSum$slope
  reg.df_2023$model_SE[i] <- modelSum$SE
  reg.df_2023$model_pval[i] <- modelSum$pvalue
  reg.df_2023$model_intercept[i] <- modelSum$intercept
  reg.df_2023$model_R2[i] <- modelSum$R2
  reg.df_2023$residual_SE[i] <- modelSum$residualSE
}

# Prepare regression df for plotting
reg_plot_2023 <- reg.df_2023 %>%
  mutate(
    pval_render = case_when(
      model_pval > 0.05 ~ paste0("p = ", formatC(model_pval, digits = 2, format = "f")),
      model_pval < 0.05 & model_pval >= 0.001 ~ paste0("p = ", formatC(model_pval, digits = 3, format = "f")),
      model_pval < 0.001 ~ "p < 0.001"
    ),
    y_label = case_when(
      variable == "CHLa" ~ "CHLa (\u03BCg L\u207B\u00B9)",
      variable == "TP" ~ "TP (mg L\u207B\u00B9)",
      variable == "TN" ~ "TN (mg L\u207B\u00B9)"
    )
  )

# ---------------------------------------------------
# 2024
reg.df_2024 <- data.frame(
  variable = c("CHLa", "TN", "TP"),
  model_slope = rep(NA, 3),
  model_SE = rep(NA, 3),
  model_pval = rep(NA, 3),
  model_intercept = rep(NA, 3),
  model_R2 = rep(NA, 3),
  residual_SE = rep(NA, 3)
)

for (i in seq_along(1:3)) {
  var <- c("CHLa", "TN", "TP")[i]
  predictor <- paste0(var,"LTM_2024")
  outcome <- paste0(var, "24")
  modelSum <- modelSummaries_2024[[paste0(outcome, "~", predictor)]]
      
  reg.df_2024$model_slope[i] <- modelSum$slope
  reg.df_2024$model_SE[i] <- modelSum$SE
  reg.df_2024$model_pval[i] <- modelSum$pvalue
  reg.df_2024$model_intercept[i] <- modelSum$intercept
  reg.df_2024$model_R2[i] <- modelSum$R2
  reg.df_2024$residual_SE[i] <- modelSum$residualSE
}

# Prepare regression df for plotting
reg_plot_2024 <- reg.df_2024 %>%
  mutate(
    pval_render = case_when(
      model_pval > 0.05 ~ paste0("p = ", formatC(model_pval, digits = 2, format = "f")),
      model_pval < 0.05 & model_pval >= 0.001 ~ paste0("p = ", formatC(model_pval, digits = 3, format = "f")),
      model_pval < 0.001 ~ "p < 0.001"
    ),
    y_label = case_when(
      variable == "CHLa" ~ "CHLa (\u03BCg L\u207B\u00B9)",
      variable == "TP" ~ "TP (mg L\u207B\u00B9)",
      variable == "TN" ~ "TN (mg L\u207B\u00B9)"
    )
  )


```

#### Write out regression stats to csv

```{r}
x <- reg.df_2023 %>%
  mutate(variable = paste0(variable, "23"))

y <- reg.df_2024 %>%
  mutate(variable = paste0(variable, "24"))

rbind(x,y)

write_csv(rbind(x,y), "data/regression_stats_annualVsLTM.csv")

```

### Merge Annual average (2023 and 2024) and long-term average values with staiton-specific trend statistics.

The newly derived annual averages and long-term averages should be merged with the station-specific trend statistics from previous water clarity analysis.
"Note that only a small proportion (possibly none) of the microcystin survey sites will have staiton-specific trend estimates for CHLa, TN or TP."

In the output dataframe, one station - 2-APP020.23 - has no station-specific trend statistics for CHLa or TN, despite having a long-term mean. Checking the metadata reveals that for CHLa, the station had *9* years of data, where our threshold was *10* for inclusion in the trend analysis. 

```{r}
station_specific_trends <- read_csv("~/Desktop/Bukaveckas_WaterLab/water-clarity-analysis/data/station_level_data/longTerm_Annual_Regression_Statistics_ALLSTATIONS.csv") %>%
    select(-model_intercept) %>%
    mutate(variable = case_when(
      grepl("NITROGEN", variable) ~ "TN",
      grepl("PHOSPHORUS", variable) ~ "TP",
      TRUE ~ variable
    )) %>%
    filter(variable %in% c("CHLa", "TN", "TP"))


x <- station_specific_trends %>%
  pivot_wider(names_from = variable, values_from = c(model_slope, model_MAD, model_pval)) %>%
  select(
    STATION_ID,
    contains("CHLa"),
    contains("TN"),
    contains("TP"),
    everything()
  )

colnames(x) <- gsub("model_", "", colnames(x))


# Merge with annual averages and LTMs
output <- main_df %>%
  left_join(x, by = c("StationId" = "STATION_ID"))

write_csv(output, "data/final_processed_dataset.csv")


```


## Data viz

#### 2023 vs LTM (through 2022)
```{r}
# 2023 vs LTM
# !------------------- CHLa -------------------------! 
plot.df_chla23 <- main_df %>%
      filter(!is.na(CHLa23))

pval_label = reg_plot_2023 %>%
  filter(variable == "CHLa") %>%
  pull(pval_render)

max_y <- max(plot.df_chla23$CHLa23, na.rm = TRUE) * 1.2
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(plot.df_chla23$CHLaLTM_2023) - 7  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.8           # Place it slightly below the upper limit of the y-axis

ggplot(plot.df_chla23, aes(x = CHLaLTM_2023, y = CHLa23)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = pval_label, size = 4.2, hjust = 1) +
      # Regression line
    geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x ="CHLa (\u03BCg L\u207B\u00B9) LTM", 
           y = "2023 Mean CHLa (\u03BCg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )
ggsave("figures/CHLa23VsLTM.jpg", width = 8, height = 6, dpi = 500)

# !------------------- TN -------------------------! 
plot.df_tn23 <- main_df %>%
      filter(!is.na(TN23))

pval_label = reg_plot_2023 %>%
  filter(variable == "TN") %>%
  pull(pval_render)


max_y <- max(plot.df_tn23$TN23, na.rm = TRUE) * 1.2
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- 1.2  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.8           # Place it slightly below the upper limit of the y-axis

ggplot(plot.df_tn23, aes(x = TNLTM_2023, y = TN23)) +
    geom_point() +
       # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = pval_label, size = 4.2, hjust = 1) +
      # Regression line
    geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x ="TN (mg L\u207B\u00B9) LTM", 
           y = "2023 Mean TN (mg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )
ggsave("figures/TN23VsLTM.jpg", width = 8, height = 6, dpi = 500)


# !------------------- TP -------------------------!
plot.df_tp23 <- main_df %>%
      filter(!is.na(TP23))
pval_label = reg_plot_2023 %>%
  filter(variable == "TP") %>%
  pull(pval_render)

max_y <- max(plot.df_tp23$TP23, na.rm = TRUE) * 1.2
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- 0.13   # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.8           # Place it slightly below the upper limit of the y-axis

ggplot(plot.df_tp23, aes(x = TPLTM_2023, y = TP23)) +
    geom_point() +
        # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = pval_label, size = 4.2, hjust = 1) +
      # Regression line
    geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x ="TP (mg L\u207B\u00B9) LTM", 
           y = "2023 Mean TP (mg L\u207B\u00B9)") +
  #scale_y_continuous(breaks = seq(0.022, .033, 0.002)) +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )
ggsave("figures/TP23VsLTM.jpg", width = 8, height = 6, dpi = 500)

```

#### 2024 vs LTM (through 2023)

```{r}
# 2024 vs LTM
# !------------------- CHLa -------------------------! 
plot.df_chla24 <- main_df %>%
      filter(!is.na(CHLa24))

pval_label = reg_plot_2023 %>%
  filter(variable == "CHLa") %>%
  pull(pval_render)

max_y <- max(plot.df_chla24$CHLa24, na.rm = TRUE)
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(plot.df_chla24$CHLaLTM_2024, na.rm = TRUE) - 5  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * .84            # Place it slightly below the upper limit of the y-axis

ggplot(plot.df_chla24, aes(x = CHLaLTM_2024, y = CHLa24)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = pval_label, size = 4.2, hjust = 1) +
      # Regression line
    geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x ="CHLa (\u03BCg L\u207B\u00B9) LTM", 
           y = "2024 Mean CHLa (\u03BCg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )
ggsave("figures/CHLa24VsLTM.jpg", width = 8, height = 6, dpi = 500)

# !------------------- TN -------------------------! 
plot.df_tn24 <- main_df %>%
      filter(!is.na(TN24))

pval_label = reg_plot_2023 %>%
  filter(variable == "TN") %>%
  pull(pval_render)

max_y <- max(plot.df_tn24$TN24, na.rm = TRUE) * 1.2
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- .6  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- .125        # Place it slightly below the upper limit of the y-axis

ggplot(plot.df_tn24, aes(x = TNLTM_2024, y = TN24)) +
    geom_point() +
       # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = pval_label, size = 4.2, hjust = 1) +
      # Regression line
    geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x ="TN (mg L\u207B\u00B9) LTM", 
           y = "2024 Mean TN (mg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )
ggsave("figures/TN24VsLTM.jpg", width = 8, height = 6, dpi = 500)


# !------------------- TP -------------------------!
plot.df_tp24 <- main_df %>%
      filter(!is.na(TP24))

pval_label = reg_plot_2023 %>%
  filter(variable == "TP") %>%
  pull(pval_render)

# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- .04   # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- .013          # Place it slightly below the upper limit of the y-axis

ggplot(plot.df_tp24, aes(x = TPLTM_2024, y = TP24)) +
    geom_point() +
        # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = pval_label, size = 4.2, hjust = 1) +
      # Regression line
    geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x ="TP (mg L\u207B\u00B9) LTM", 
           y = "2024 Mean TP (mg L\u207B\u00B9)") +
  #scale_y_continuous(breaks = seq(0.022, .033, 0.002)) +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )
ggsave("figures/TP24VsLTM.jpg", width = 8, height = 6, dpi = 500)

```
