---
title: "Part 2: Station Average Regressions for Variable Pairs"
author: "Andrew Cameron"
date: "2025-2-23"
output: html_document
---

CHLa vs TN, CHLa vs. TP, TN vs. TP, MC vs CHLa, MC vs. TN and MC vs. CHLa (with model lines shown, confidence intervals as shading, p values and R2 values).  These are all based on ordinary least squares regression.

```{r}
source("C:/Users/andre/my_functions.R")

library(tidyverse)
library(ggplot2)
data<- openxlsx::read.xlsx("data/received/Integrated Dataset 2023-24 Reservoir Microcystin Survey.xlsx", sheet = 4, startRow = 2) %>%
  select(1:10) %>%
  filter(!row_number() %in% c(277, 278)) %>%
  mutate(TP = as.numeric(TP)) %>%
  mutate(Date = convertExcelDateTime(., "Date")) %>%
  mutate(month = factor(month(Date), levels = 1:12, labels = month.abb))


other_columns <- data %>% 
  select(StationID, StationName, RegionalOffice)   # to be joined back later. Can't be retained in grouping/summarizing because lots of NA for StationName and AdminRegion.
```


## Output 1: Regressions of variable pair station-year means 

CHLa vs TN, 
CHLa vs. TP,'
TN vs. TP, 
MC vs CHLa,
MC vs. TN 
and MC vs. CHLa

```{r}
data_avg <- data %>%
  mutate(year = year(Date)) %>%
  group_by(StationID, year) %>%
  summarise(CHLa = mean(CHLa, na.rm = TRUE),
            TN = mean(TN, na.rm = TRUE),
            TP = mean(TP, na.rm = TRUE),
            MC =  mean(MC, na.rm = TRUE),
            .groups = "drop")



# Run models for each variable pair
modSums_output <- list()  
  
  for (var_pair in list(
    c("CHLa", "TP"), c("CHLa", "TN"), c("TN", "TP"), 
    c("MC", "CHLa"), c("MC", "TN"), c("MC", "TP"))
    ) {
    x <- var_pair[2]
    y <- var_pair[1]
    list_id <- paste(y, "vs", x)
    
    model_data <- data_avg %>%
      filter(!is.na(!!sym(x)) & !is.na(!!sym(y)))
    
    formula <- as.formula(paste(y, "~", x))
    
    # Run mblm model
    model <- lm(formula, data = model_data)
    mod.sum <- summary(model)
    
    # Store results (directly, without creating the combined station-variable name)
    modelSummary <- list(
            variable = var,
            slope = mod.sum$coefficients[2, 1],
            SE = mod.sum$coefficients[2, 2],
            intercept = mod.sum$coefficients[1, 1],
            R2 = mod.sum$r.squared,
            residualSE = mod.sum$sigma,
            pvalue = mod.sum$coefficients[2, 4]
          )
    
    modSums_output[[list_id]] <- modelSummary
  }

```


```{r `regression stats`}
# Create single data frame with regression statistics
reg.df_output <- data.frame(
  model = c("CHLa vs TP", "CHLa vs TN", "TN vs TP", "MC vs CHLa", "MC vs TN", "MC vs TP"),
  model_slope = rep(NA, 6),
  model_SE = rep(NA, 6),
  model_pval = rep(NA, 6),
  model_intercept = rep(NA, 6),
  model_R2 = rep(NA, 6),
  residual_SE = rep(NA, 6)
)

for (i in seq_along(reg.df_output$model)) {
  modelSum <- modSums_output[[reg.df_output$model[i]]]
  
  reg.df_output$model_slope[i] <- modelSum$slope
  reg.df_output$model_SE[i] <- modelSum$SE
  reg.df_output$model_pval[i] <- modelSum$pvalue
  reg.df_output$model_intercept[i] <- modelSum$intercept
  reg.df_output$model_R2[i] <- modelSum$R2
  reg.df_output$residual_SE[i] <- modelSum$residualSE
}

# Write results to CSV
write_csv(reg.df_output, "data/derived/reg_stats_varPairs_20250223.csv")

# Prepare regression data frame for plotting
reg_plot <- reg.df_output %>%
  mutate(
    pval_render = case_when(
      model_pval > 0.05 ~ paste0("p = ", formatC(model_pval, digits = 2, format = "f")),
      model_pval < 0.05 & model_pval >= 0.001 ~ paste0("p = ", formatC(model_pval, digits = 3, format = "f")),
      model_pval < 0.001 ~ "p < 0.001"
    ),
    y_label = case_when(
      grepl("TN vs", model) ~ "TN (mg L\u207B\u00B9)",
      grepl("CHLa vs", model) ~ "CHLa (\u03BCg L\u207B\u00B9)",
      grepl("MC vs", model) ~ "MC (\u03BCg L\u207B\u00B9)"
    )
  )

```

### Data viz

```{r}
# !------------------- CHLa vs TP -------------------------! 
chlaVsTP <- data_avg %>%
      filter(!is.na(CHLa) & !is.na(TP))
reg_chlaVsTP <- reg_plot %>%
  filter(model == "CHLa vs TP")

max_y <- max(chlaVsTP$CHLa, na.rm = TRUE) * 1.05
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- 0.128 
  pval_y_pos <- 30         

 p <- ggplot(chlaVsTP, aes(x = TP, y = CHLa)) +
    geom_point(alpha = 0.7) +
      # Add p-value inside the plot, positioning dynamically
    annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = paste0(reg_chlaVsTP$pval_render, 
                              "\n R2 = ",
                              round(reg_chlaVsTP$model_R2, 3)
                              ),
                              size = 4.2, hjust = 1) +
      # Regression line
     geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x = "TP (mg L\u207B\u00B9)", 
           y = "CHLa (\u03BCg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )

#ggsave("figures/CHLa_vs_TP_stationYrAvg.jpg", width = 8, height = 6, dpi = 500)
saveRDS(p, "data/R_data/CHLa_vs_TP_stationYrAvg.rds")

# !------------------- CHLa vs TN -------------------------! 
chlaVsTN <- data_avg %>%
      filter(!is.na(CHLa) & !is.na(TN))
reg_chlaVsTN <- reg_plot %>%
  filter(model == "CHLa vs TN")

max_y <- max(chlaVsTN$CHLa, na.rm = TRUE) * 1.05
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- 1.5
  pval_y_pos <- 25        

p <- ggplot(chlaVsTN, aes(x = TN, y = CHLa)) +
    geom_point(alpha = 0.7) +
      # Add p-value inside the plot, positioning dynamically
    annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = paste0(reg_chlaVsTN$pval_render, 
                              "\n R2 = ",
                              round(reg_chlaVsTN$model_R2, 3)
                              ),
                              size = 4.2, hjust = 1) +
      # Regression line
      geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x = "TN (mg L\u207B\u00B9)", 
           y = "CHLa (\u03BCg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )

#ggsave("figures/CHLa_vs_TN_stationYrAvg.jpg", width = 8, height = 6, dpi = 500)
saveRDS(p, "data/R_data/CHLa_vs_TN_stationYrAvg.rds")

# !------------------- TN vs TP -------------------------! 
TNvsTP <- data_avg %>%
      filter(!is.na(TP) & !is.na(TN))
reg_TNvsTP <- reg_plot %>%
  filter(model == "TN vs TP")

max_y <- max(TNvsTP$TN, na.rm = TRUE)
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- 0.13
  pval_y_pos <- 0.5          

p <- ggplot(TNvsTP, aes(x = TP, y = TN)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
    annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = paste0(reg_TNvsTP$pval_render, 
                              "\n R2 = ",
                              round(reg_TNvsTP$model_R2, 3)),
                              size = 4.2, hjust = 1) +
      # Regression line
    geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           y = "TN (mg L\u207B\u00B9)", 
           x = "TP (mg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )

#ggsave("figures/TN_vs_TP_stationYrAvg.jpg", width = 8, height = 6, dpi = 500)
saveRDS(p, "data/R_data/TN_vs_TP_stationYrAvg.rds")

# !------------------- MC vs CHLa -------------------------! 
MCvsCHLa <- data_avg %>%
      filter(!is.na(MC) & !is.na(CHLa))
reg_MCvsCHLA<- reg_plot %>%
  filter(model == "MC vs CHLa")

max_y <- max(MCvsCHLa$MC, na.rm = TRUE) * 1.05
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- 50
  pval_y_pos <- 0.68    

p <- ggplot(MCvsCHLa, aes(x = CHLa, y = MC)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = paste0(reg_MCvsCHLA$pval_render),
                              size = 4.2, hjust = 1) +
      # Regression line
    geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x = "CHLa (\u03BCg L\u207B\u00B9)",
           y = "MC (\u03BCg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )

#ggsave("figures/MC_vs_CHLa_stationYrAvg.jpg", width = 8, height = 6, dpi = 500)
saveRDS(p, "data/R_data/MC_vs_CHLa_stationYrAvg.rds")

# !------------------- MC vs TN -------------------------! 
MCvsTN <- data_avg %>%
      filter(!is.na(MC) & !is.na(TN))
reg_MCvsTN <- reg_plot %>%
  filter(model == "MC vs TN")

max_y <- max(MCvsTN$MC, na.rm = TRUE) * 1.05
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- 1
  pval_y_pos <- 0.68         
  
p <- ggplot(MCvsTN, aes(x = TN, y = MC)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = paste0(reg_MCvsTN$pval_render), 
                              size = 4.2, hjust = 1) +
      # Regression line
    geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x = "TN (mg L\u207B\u00B9)",
           y = "MC (\u03BCg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )

#ggsave("figures/MC_vs_TN_stationYrAvg.jpg", width = 8, height = 6, dpi = 500)
saveRDS(p, "data/R_data/MC_vs_TN_stationYrAvg.rds")


# !------------------- MC vs TP -------------------------! 
MCvsTP <- data_avg %>%
      filter(!is.na(MC) & !is.na(TP))
reg_MCvsTP <- reg_plot %>%
  filter(model == "MC vs TP")

max_y <- max(MCvsTP$MC, na.rm = TRUE) * 1.05
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- 0.1
  pval_y_pos <- 0.68            
  
p <- ggplot(MCvsCHLa, aes(x = TP, y = MC)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = paste0(reg_MCvsTP$pval_render),
                              size = 4.2, hjust = 1) +
      # Regression line
    geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x = "TP (mg L\u207B\u00B9)",
           y = "MC (\u03BCg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )

#ggsave("figures/MC_vs_TP_stationYrAvg.jpg", width = 8, height = 6, dpi = 500)
saveRDS(p, "data/R_data/MC_vs_TP_stationYrAvg.rds")



```


## Output 2: Regressions of variable pairs, indiv obs 

CHLa vs TN, 
CHLa vs. TP,'
TN vs. TP, 
MC vs CHLa,
MC vs. TN 
and MC vs. CHLa

```{r}
# Run models for each variable pair
modSums_output2 <- list()  
  
  for (var_pair in list(
    c("CHLa", "TP"), c("CHLa", "TN"), c("TN", "TP"), 
    c("MC", "CHLa"), c("MC", "TN"), c("MC", "TP"))
    ) {
    x <- var_pair[2]
    y <- var_pair[1]
    list_id <- paste(y, "vs", x)
    
    model_data <- data %>%
      filter(!is.na(!!sym(x)) & !is.na(!!sym(y)))
    
    formula <- as.formula(paste(y, "~", x))
    
    # Run mblm model
    model <- lm(formula, data = model_data)
    mod.sum <- summary(model)
    
    # Store results (directly, without creating the combined station-variable name)
    modelSummary <- list(
            variable = var,
            slope = mod.sum$coefficients[2, 1],
            SE = mod.sum$coefficients[2, 2],
            intercept = mod.sum$coefficients[1, 1],
            R2 = mod.sum$r.squared,
            residualSE = mod.sum$sigma,
            pvalue = mod.sum$coefficients[2, 4]
          )
    
    modSums_output2[[list_id]] <- modelSummary
  }

```


```{r `regression stats`}
# Create single data frame with regression statistics
reg.df_output2 <- data.frame(
  model = c("CHLa vs TP", "CHLa vs TN", "TN vs TP", "MC vs CHLa", "MC vs TN", "MC vs TP"),
  model_slope = rep(NA, 6),
  model_SE = rep(NA, 6),
  model_pval = rep(NA, 6),
  model_intercept = rep(NA, 6),
  model_R2 = rep(NA, 6),
  residual_SE = rep(NA, 6)
)

for (i in seq_along(reg.df_output2$model)) {
  modelSum <- modSums_output2[[reg.df_output2$model[i]]]
  
  reg.df_output2$model_slope[i] <- modelSum$slope
  reg.df_output2$model_SE[i] <- modelSum$SE
  reg.df_output2$model_pval[i] <- modelSum$pvalue
  reg.df_output2$model_intercept[i] <- modelSum$intercept
  reg.df_output2$model_R2[i] <- modelSum$R2
  reg.df_output2$residual_SE[i] <- modelSum$residualSE
}

# Write results to CSV
write_csv(reg.df_output2, "data/derived/regressions_varPairs_indivObs_20250223.csv")

# Prepare regression data frame for plotting
reg_plot2 <- reg.df_output2 %>%
  mutate(
    pval_render = case_when(
      model_pval > 0.05 ~ paste0("p = ", formatC(model_pval, digits = 2, format = "f")),
      model_pval < 0.05 & model_pval >= 0.001 ~ paste0("p = ", formatC(model_pval, digits = 3, format = "f")),
      model_pval < 0.001 ~ "p < 0.001"
    ),
    y_label = case_when(
      grepl("TN vs", model) ~ "TN (mg L\u207B\u00B9)",
      grepl("CHLa vs", model) ~ "CHLa (\u03BCg L\u207B\u00B9)",
      grepl("MC vs", model) ~ "MC (\u03BCg L\u207B\u00B9)"
    )
  )

```

### Data viz

```{r}
# !------------------- CHLa vs TP -------------------------! 
chlaVsTP <- data %>%
      filter(!is.na(CHLa) & !is.na(TP))
reg_chlaVsTP <- reg_plot2 %>%
  filter(model == "CHLa vs TP")

max_y <- max(chlaVsTP$CHLa, na.rm = TRUE) * 1.05
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(chlaVsTP$TP) - .03  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.95           # Place it slightly below the upper limit of the y-axis

 p <- ggplot(chlaVsTP, aes(x = TP, y = CHLa)) +
    geom_point(alpha = 0.7) +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = paste0(reg_chlaVsTP$pval_render, 
                              "\n R2 = ",
                              round(reg_chlaVsTP$model_R2, 3)
                              ),
                              size = 4.2, hjust = 1) +
      # Regression line
     geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x = "TP (mg L\u207B\u00B9)", 
           y = "CHLa (\u03BCg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )

ggsave("figures/CHLa_vs_TP_indivObs.jpg", width = 8, height = 6, dpi = 500)


# !------------------- CHLa vs TN -------------------------! 
chlaVsTN <- data %>%
      filter(!is.na(CHLa) & !is.na(TN))
reg_chlaVsTN <- reg_plot2 %>%
  filter(model == "CHLa vs TN")

max_y <- max(chlaVsTN$CHLa, na.rm = TRUE) * 1.05
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- 1.32
  pval_y_pos <- max_y * 0.95          

ggplot(chlaVsTN, aes(x = TN, y = CHLa)) +
    geom_point(alpha = 0.7) +
      # Add p-value inside the plot, positioning dynamically
    annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = paste0(reg_chlaVsTN$pval_render, 
                              "\n R2 = ",
                              round(reg_chlaVsTN$model_R2, 3)
                              ),
                              size = 4.2, hjust = 1) +
      # Regression line
      geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x = "TN (mg L\u207B\u00B9)", 
           y = "CHLa (\u03BCg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )

ggsave("figures/CHLa_vs_TN_indivObs.jpg", width = 8, height = 6, dpi = 500)


# !------------------- TN vs TP -------------------------! 
TNvsTP <- data %>%
      filter(!is.na(TP) & !is.na(TN))
reg_TNvsTP <- reg_plot2 %>%
  filter(model == "TN vs TP")

max_y <- max(TNvsTP$TN, na.rm = TRUE) * 1.05
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- 0.13 
  pval_y_pos <- max_y * 0.95           

ggplot(TNvsTP, aes(x = TP, y = TN)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
    annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = paste0(reg_TNvsTP$pval_render, 
                              "\n R2 = ",
                              round(reg_TNvsTP$model_R2, 3)),
                              size = 4.2, hjust = 1) +
      # Regression line
    geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           y = "TN (mg L\u207B\u00B9)", 
           x = "TP (mg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )

ggsave("figures/TN_vs_TP_indivObs.jpg", width = 8, height = 6, dpi = 500)


# !------------------- MC vs CHLa -------------------------! 
MCvsCHLa <- data %>%
      filter(!is.na(MC) & !is.na(CHLa))
reg_MCvsCHLA<- reg_plot2 %>%
  filter(model == "MC vs CHLa")

max_y <- max(MCvsCHLa$MC, na.rm = TRUE) * 1.05
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- 50  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.95           # Place it slightly below the upper limit of the y-axis

p <- ggplot(MCvsCHLa, aes(x = CHLa, y = MC)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = paste0(reg_MCvsCHLA$pval_render, 
                              "\n R2 = ",
                              round(reg_MCvsCHLA$model_R2, 3)
                              ,
                              size = 4.2, hjust = 1)) +
      # Regression line
    geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x = "CHLa (\u03BCg L\u207B\u00B9)",
           y = "MC (\u03BCg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )

ggsave("figures/MC_vs_CHLa_indivObs.jpg", width = 8, height = 6, dpi = 500)


# !------------------- MC vs TN -------------------------! 
MCvsTN <- data %>%
      filter(!is.na(MC) & !is.na(TN))
reg_MCvsTN <- reg_plot2 %>%
  filter(model == "MC vs TN")

max_y <- max(MCvsTN$MC, na.rm = TRUE) * 1.05
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- 1
  pval_y_pos <- max_y * 0.95           
  
p <- ggplot(MCvsTN, aes(x = TN, y = MC)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = paste0(reg_MCvsTN$pval_render, 
                              "\n R2 = ",
                              round(reg_MCvsTN$model_R2, 3)),
                              size = 4.2, hjust = 1) +
      # Regression line
    geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x = "TN (mg L\u207B\u00B9)",
           y = "MC (\u03BCg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )

ggsave("figures/MC_vs_TN_indivObs.jpg", width = 8, height = 6, dpi = 500)


# !------------------- MC vs TP -------------------------! 
MCvsTP <- data %>%
      filter(!is.na(MC) & !is.na(TP))
reg_MCvsTP <- reg_plot2 %>%
  filter(model == "MC vs TP")

max_y <- max(MCvsTP$MC, na.rm = TRUE) * 1.05
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- 0.1
  pval_y_pos <- max_y * 0.95           
  
p <- ggplot(MCvsCHLa, aes(x = TP, y = MC)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = paste0(reg_MCvsTP$pval_render, 
                              "\n R2 = ",
                              round(reg_MCvsTP$model_R2, 3))
                              ,
                              size = 4.2, hjust = 1) +
      # Regression line
    geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x = "TP (mg L\u207B\u00B9)",
           y = "MC (\u03BCg L\u207B\u00B9)") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )

ggsave("figures/MC_vs_TP_indivObs.jpg", width = 8, height = 6, dpi = 500)



```






































## output from initial analysis, no longer needed


Regressions of station average values for 2023 vs. 2024


```{r}
#Derive average values for each year
data_avg <- data %>%
  mutate(year = year(Date)) %>%
  group_by(StationID, Depth, year) %>%
  summarise(CHLa = mean(CHLa, na.rm = TRUE),
            TN = mean(TN, na.rm = TRUE),
            TP = mean(TP, na.rm = TRUE),
            MC =  mean(MC, na.rm = TRUE),
            .groups = "drop")

#Array results by column
data_avg_wide <- data_avg %>%
  pivot_wider(names_from = year, values_from = c(CHLa, TN, TP))

desired_names <- c("CHLa23", "CHLa24", "TN23", "TN24", "TP23", "TP24")
names(data_avg_wide)[3:8] <- desired_names

#Bring back StationName and AdminRegion
stationName <- other_columns %>%
  group_by(StationID) %>%
  summarise(
    StationName = coalesce(first(StationName[!is.na(StationName)]), first(StationName))
  ) %>%
  ungroup()

RegionalOffice <- other_columns %>%
  group_by(StationID) %>%
  summarise(
    RegionalOffice = coalesce(first(RegionalOffice[!is.na(RegionalOffice)]), first(RegionalOffice))
  ) %>%
  ungroup()

data_avg_wide <- data_avg_wide %>%
  left_join(stationName, by = "StationID") %>%
  left_join(RegionalOffice, by = "StationID") %>%
  relocate(StationName, RegionalOffice, .before = Depth)

write_csv(data_avg_wide, "data/station_avg_by_year.csv")

```

How many stations are actually sampled in both years?

```{r}
both_chla <- data_avg_wide %>%
  filter(!is.na(CHLa23) & !is.na(CHLa24))

both_tn <- data_avg_wide %>%
  filter(!is.na(TN23) & !is.na(TN24))

both_tp <- data_avg_wide %>%
  filter(!is.na(TP23) & !is.na(TP24))

print(paste("CHLa: ", nrow(both_chla)))
print(paste("TN: ", nrow(both_tn)))
print(paste("TP: ", nrow(both_tp)))

```

```{r, warning = FALSE}

modelSummaries <- list()  
  
  for (var in c("CHLa", "TN", "TP")) {
  
    model_data <- data %>%
      filter(!is.na(!!sym(paste0(var, "23"))) & !is.na(!!sym(paste0(var, "24"))))
    
    modeling_cols <- c(paste0(var, "23"), paste0(var, "24"))
    formula <- as.formula(paste0(modeling_cols[2], " ~ ", modeling_cols[1]))
    
    # Run mblm model
    model <- lm(formula, data = model_data)
    mod.sum <- summary(model)
    
  
    # Store results (directly, without creating the combined station-variable name)
          modelSummary <- list(
            variable = var,
            slope = mod.sum$coefficients[2, 1],
            SE = mod.sum$coefficients[2, 2],
            intercept = mod.sum$coefficients[1, 1],
            R2 = mod.sum$r.squared,
            residualSE = mod.sum$sigma,
            pvalue = mod.sum$coefficients[2, 4]
          )
          
    modelSummaries[[var]] <- modelSummary
  }
  
```

#### Extract Regression Statistics into single table

```{r `regression stats`}
# Extract Regression Statistics into Single Table
reg.df <- data.frame(
  variable = c("CHLa", "TN", "TP"),
  model_slope = rep(NA, 3),
  model_SE = rep(NA, 3),
  model_pval = rep(NA, 3),
  model_intercept = rep(NA, 3),
  model_R2 = rep(NA, 3),
  residual_SE = rep(NA, 3)
)

for (i in seq_along(1:3)) {
  var <- c("CHLa", "TN", "TP")[i]
  modelSum <- modelSummaries[[var]]
      
  reg.df$model_slope[i] <- modelSum$slope
  reg.df$model_SE[i] <- modelSum$SE
  reg.df$model_pval[i] <- modelSum$pvalue
  reg.df$model_intercept[i] <- modelSum$intercept
  reg.df$model_R2[i] <- modelSum$R2
  reg.df$residual_SE[i] <- modelSum$residualSE
}

# Write results to csv
write_csv(reg.df, "data/regression_stats_23vs24_byVariable.csv")


# Prepare regression df for plotting
reg_plot <- reg.df %>%
  mutate(
    pval_render = case_when(
      model_pval > 0.05 ~ paste0("p = ", formatC(model_pval, digits = 2, format = "f")),
      model_pval < 0.05 & model_pval >= 0.001 ~ paste0("p = ", formatC(model_pval, digits = 3, format = "f")),
      model_pval < 0.001 ~ "p < 0.001"
    ),
    y_label = case_when(
      variable == "CHLa" ~ "CHLa (\u03BCg L\u207B\u00B9)",
      variable == "TP" ~ "TP (mg L\u207B\u00B9)",
      variable == "TN" ~ "TN (mg L\u207B\u00B9)"
    )
  )

```

#### Data viz

```{r}
# !------------------- CHLa -------------------------! 
data_chla <- data_avg_wide %>%
      filter(!is.na(CHLa23) & !is.na(CHLa24))
reg_chla <- reg_plot %>%
  filter(variable == "CHLa")   
max_y <- max(data_chla$CHLa23, na.rm = TRUE) * 1.2
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(data_chla$CHLa23) - 12  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.95           # Place it slightly below the upper limit of the y-axis

ggplot(data_chla, aes(x = CHLa23, y = CHLa24)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = reg_chla$pval_render, size = 4.2, hjust = 1) +
      # Regression line
          geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x ="CHLa (\u03BCg L\u207B\u00B9) 2023", 
           y = "CHLa (\u03BCg L\u207B\u00B9) 2024") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )
ggsave("figures/CHLa_23vs24.jpg", width = 8, height = 6, dpi = 500)

# !------------------- TN -------------------------! 
data_TN <- data_avg_wide %>%
      filter(!is.na(TN23) & !is.na(TN24))
reg_TN <- reg_plot %>%
  filter(variable == "TN")
max_y <- max(data_TN$TN23, na.rm = TRUE) * 1.2
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(data_TN$TN23) -.05  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.5           # Place it slightly below the upper limit of the y-axis

ggplot(data_TN, aes(x = TN23, y = TN24)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = reg_TN$pval_render, size = 4.2, hjust = 1) +
      # Regression line
          geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x ="TN (mg L\u207B\u00B9) 2023", 
           y = "TN (mg L\u207B\u00B9) 2024") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )
ggsave("figures/TN_23vs24.jpg", width = 8, height = 6, dpi = 500)


# !------------------- TP -------------------------!
data_TP <- data_avg_wide %>%
      filter(!is.na(TP23) & !is.na(TP24))
reg_TP <- reg_plot %>%
  filter(variable == "TP")

max_y <- max(data_TP$TP24, na.rm = TRUE) * 1.2
# Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(data_TP$TP23) - .01   # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.8           # Place it slightly below the upper limit of the y-axis

ggplot(data_TP, aes(x = TP23, y = TP24)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = reg_TP$pval_render, size = 4.2, hjust = 1) +
      # Regression line
          geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = .5 ) +
      labs(title = NULL,
           x ="TP (mg L\u207B\u00B9) 2023", 
           y = "TP (mg L\u207B\u00B9) 2024") +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 14),        # Adjust axis title size if needed
        axis.text = element_text(size = 10, color = "black"),         # Adjust axis text size
        plot.margin = unit(c(.2, .2, .2, .2), "cm")
      )

ggsave("figures/TP_23vs24.jpg", width = 8, height = 6, dpi = 500)


```
